---
title: "Extended Data Figure 5"
author: "Regan"
date: "16/05/2020"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

This script will generate the scRNA-seq data figures in supplementary figure 3 from the dataset that has been pre-processed with scripts 1-6.

# Extended Data Fig 5 Composition plots - cluster by stage
- using the Monocle3-derived clusters

Load the sce and set the directory
```{r}
rm(list=ls())

sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

dirC <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5a/"
```

Generate the area plot 
```{r}
library(tibble)
library(ggplot2)
library(plyr)
library(RColorBrewer)

CT <- "MCd"
sce_c <- sce[,sce$cell_type == CT]

# set HC to 0
sce_c$stage <- factor(sce_c$stage, levels = c("HC","0", "1", "2", "3", "10", "21"))
sce_c$stage[sce_c$stage == "HC"] <- 0

# set the x and y values for the plot
y <- sce_c$Clust_final_names
x <- droplevels(factor(sce_c$stage))

# create the df containing the x and y values
df <- as.data.frame(table(y, x))
colnames(df) <- c("y", "x", "Freq")

# create data
time <- as.numeric(levels(df$x)[df$x]) # x Axis
value <- as.numeric(df$Freq)             # y Axis
group <- df$y       # group, one shape per group
data <- data.frame(time, value, group)

# Compute percentages with dplyr
library(dplyr)

data <- data  %>%
  group_by(time, group) %>%
  summarise(n = sum(value)) %>%
  mutate(percentage = n / sum(n))

# Plot
ggplot(data, aes(x=time, y=percentage, fill=group)) + 
  geom_area(alpha=0.7 , size=0.5, colour="black") + ylab("Fraction of Cells Per Cluster") + scale_fill_manual(values=c("lightblue", brewer.pal(11,"RdYlGn"),"#003319","black", "grey50", "grey90"), drop=F)+ scale_x_continuous(name ="Day Post-SCI", breaks=c(0,1,2, 3, 10, 21), labels=c("0"="HC","1"="1","2"="2", "3"="3", "10"="10", "21"="21"))

ggsave(paste0(dirC, CT, "-Area Plot clust vs stage.pdf"), height=5, width=7)
```

# Extended Data Fig 5b Pearson's Chi-squared test for independence, plotting the standardized residuals

- Null hypothesis: cluster membership is independent of the post-SCI timepoint
```{r}
sce_C <- sce[,sce$stage != "HC"]

# generate a contingency table
tbl <- table(droplevels(sce_C$stage), droplevels(sce_C$Clust_final_names))

# test for independence using Pearson's Chi-square
chisq <- chisq.test(tbl)

# convert the z-scores (standardized residuals) to chi square values
chisq.val <- chisq$stdres * chisq$stdres

# convert the chi square values to p values
chi.p <- pchisq(chisq.val, df=1, lower.tail=FALSE) 

# correct for multiple testing:
# compare the chi square values to the bonferonni adjusted p value
bf.p <- (0.05/(nrow(tbl)*ncol(tbl)))
```

Plot the chi-squared residuals
```{r}
library(corrplot)
pdf(paste0("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5.pdf"))
corrplot(chisq$stdres, p.mat=chi.p, is.cor = FALSE, sig.level = bf.p, tl.col = "black")
dev.off()
```

Save the chi-squared results for Extended Data Table 2
```{r}
# write the parameters
write.csv(x=cbind(chisq$parameter, chisq$p.value, chisq$statistic, chisq$method), file= "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Stats/Chi_parameters.csv")

# write the outputs
write.csv(x=rbind(chisq$observed, chisq$expected, chisq$residuals, chisq$stdres), file= "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Stats/Chi_all.csv")

# write the converted chisquare values and p values
write.csv(x=rbind(chisq.val, chi.p), file= "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Stats/chisq and p values.csv")
```


# Extended Data Fig 5c, g, h Boxplots of a single gene over all clusters 
- using the Monocle3-derived clusters

Load the sce, reorder cluster names, and set the directory
```{r}
rm(list=ls())
sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

dir <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5c, g, h/"
```

Set the genes of interest to be plotted
```{r}
genes <- c("Fabp5", "Csf1r", "Mki67", "Trem2", "Ccr2")
```

Generate the box plot
```{r}
# set the clusters
x <- sce$Clust_final_names

# generate plots for each gene
for (i in genes) {
  
  # extract the log10-transformed counts for the gene of interest
  logs <- as.numeric(logcounts(sce[i]))
  df <- data.frame(Logs=logs, x=x)
  
  #plot
  ggplot(data=df, aes(y=Logs, x=x, fill=x)) + geom_boxplot() + ggtitle(i) + theme(plot.title=element_text(hjust=0.5), axis.text.x=element_text(angle=45, hjust=1), legend.position="none") + ylab("Logcounts") + xlab("Cluster") + scale_fill_manual(values=c("lightblue", brewer.pal(11,"RdYlGn"),"#003319","black", "grey50", "grey50"))  
  
  # save and size
  ggsave(paste0(dir,"byClust -", i, ".pdf"), scale=1, height=2, width=6)
}
```

# Figure 3d UMAP displaying DAM1 genes vs DAM2 genes 
- using Seurat to plot the expression of two genes on one UMAP

Load the sce and set the directory
```{r}
rm(list=ls())
sce_u <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

dirU <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5d -"
```

Load the function for preparing gene lists from source files
```{r}
source("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/scRNAseq Workflow/Regan et al 202X/Sources/Fxn - Prep Genes.R")
```

Establish the DAM gene lists (from Keren-Shaul et al., 2017; Figure 6)
```{r}
# DAM phase 1 (Trem independent)
DAM1 <- c("Tyrobp, Ctsb, Ctsd, Apoe, B2m, Fth1, Lyz2") 
names(DAM1) <- "DAM1"

# DAM phase 2 (Trem dependent)
DAM2 <- c("Trem2, Axl, Cst7, Ctsl, Lpl, Cd9, Csf1, Ccl6, Itgax, Clec7a, Lilrb4a, Timp2")  
names(DAM2) <- "DAM2"

# separate the string
DAM1 <- prepG(DAM1, sep=", ", match=NULL)

DAM2 <- prepG(DAM2, sep=", ", match=NULL)
```

Calculate number of DAM genes expressed per cell
```{r}
library(scater)

#sce_u <- calculateQCMetrics(sce_u,feature_controls= list(DAM1=DAM1, DAM2=DAM2))
sce_u$logMeanDAM2 <- colMeans(logcounts(sce_u[DAM2]))
sce_u$logMeanDAM1 <- colMeans(logcounts(sce_u[DAM1]))
```

Plot the UMAP
```{r}
library(Seurat)

# convert the sce to a seurat object
seur <- as.Seurat(sce_u[,sce_u$cell_type != "NP"])#[,sce_u$partition ==1])

# select the metrics to plot
x <- "logMeanDAM1" #total_features_by_counts_DAM1 log10_total_counts_DAM1
y <- "logMeanDAM2"

# generate the UMAPs
list <- FeaturePlot(object=seur,
                    features=c(x, y), 
                    cols=c("grey60", "purple", "green"), 
                    reduction="UMAP",
                    blend=TRUE,
                    combine=F,
                    blend.threshold=0,
                    pt.size=0.5)


# convert the UMAP dots to a single image (easier handling in Illustrator) and save
AugmentPlot(plot=list[[3]]+NoLegend(), dpi=600)  
ggsave(paste0(dirU, x, " vs ", y,", FIG6.pdf"), scale=0.75, width=7, height=8,units=c("in", "cm", "mm"), dpi=300, limitsize=TRUE)

# plot and save the legend
list[[4]]+NoLegend()
ggsave(paste0(dirU, x, "vs", y, "legend.pdf"), scale=0.75, width=8, height=8,units=c("in", "cm", "mm"), dpi=300, limitsize=TRUE)
```

# Extended Data Figure 5e UMAP displaying Fabp5 vs Trem2 expression
- using Seurat to plot the expression of two genes on one UMAP

Load the sce and set the directory
```{r}
rm(list=ls())
sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

# remove neutrophils
NP <- sce@int_colData@listData[["reducedDims"]]@listData[["UMAP"]][,1] < -5
sce_u <- sce[, !NP]

dirU <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5e-"
```

Plot the UMAP
```{r}
library(Seurat)
library(ggplot2)

# convert the sce to a seurat object
seur <- as.Seurat(sce_u, data = "logcounts", assay="logcounts")

# select the genes to plot
x <- "Trem2" 
y <- "Fabp5"

# generate the UMAPs
list <- FeaturePlot(object=seur,
                    features=c(x, y), 
                    cols=c("grey60", "purple", "green"), 
                    reduction="UMAP",
                    blend=TRUE,
                    combine=F,
                    blend.threshold=0,
                    pt.size=1)

# convert the UMAP dots to a single image (easier handling in Illustrator)
AugmentPlot(plot=list[[3]]+NoLegend() + labs(title=NULL), dpi=600)  

# plot and save the UMAP without a legend (to be plotted separtely)
ggsave(paste0(dirU, x, " vs ", y, ".pdf"), scale=0.75, width=7, height=8,units=c("in", "cm", "mm"), dpi=300, limitsize=TRUE)

# plot and save the legend
list[[4]]+NoLegend()
ggsave(paste0(dirU, x, "vs", y, "legend.pdf"), scale=0.75, width=8, height=8,units=c("in", "cm", "mm"), dpi=300, limitsize=TRUE)
```

# Extended Data Fig 5f Heatmap of cell cycling genes

Load the sce and set the directory
```{r}
rm(list=ls())
sce_hm_all <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")
dirH <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5f/"

# the plotHeatmap function requires unique column names, rename any duplicated columns in the sce object
colnames(sce_hm_all) <- make.unique(colnames(sce_hm_all), sep="X")
```

Set the genes for the heatmap
```{r}
#  read in a list of cell cycle markers, from Tirosh et al., 2015
cc.genes <- readLines(con="/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/cell_cycle_vignette_files/regev_lab_cell_cycle_genes.txt")
cc.genes <- stringr::str_to_title(cc.genes)

# s phase vs g2-m phase genes for label reference
s.genes <- cc.genes[1:43]
g2m.genes <- cc.genes[44:97]

# load the function for preparing gene lists from source file
source("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/scRNAseq Workflow/Regan et al 202X/Sources/Fxn - Prep Genes.R")

# format the list of genes gfor the plotHeatmap function
input <- prepG(cc.genes, sep=",")

# remove any genes not in the sce
input <- match(input, rownames(sce_hm_all))
input <- input[!is.na(input)]
```

Generate the heatmaps for each cell type of interest 
```{r}
library(scater)

# set the cell type
CT <- "MCd" #"MG"
sce_hm <- sce_hm_all[,sce_hm_all$cell_type == CT]

# plot and save the heatmap
plotHeatmap(sce_hm, features=input, exprs_values="logcounts", cellheight=4, cellwidth=0.03,
            zlim=NULL, center=FALSE, symmetric=FALSE, cluster_cols=FALSE,cluster_rows=F,
            colour_columns_by=c("stage","Clust_final_names"), columns=order(sce_hm$Clust_final_names), main= paste("S-G2-M Phase Genes"), fontsize_row=5, fontsize_col=0.2, treeheight_col=40, file=paste0(dirH, "S-G2-M Phase Genes stage", CT, ".pdf"))  
```

# Extended Data Fig 5i Heatmap of MG(A2) cluster DEGs
- DEGs were generated in script 6 from the Monocle3-derived clusters

Load the sce and markers, and set the directory
```{r}
rm(list=ls())

sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

markers <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Monocle/k=16 cluster markers/subcluster markers.RDS")

dirH <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Extended Data Fig 5i - MG(A2).pdf"
```

Extract the Cremato cells from cluster 2-MG
```{r}
sce_c <- sce[,sce$Clust_final_names == "MG(A2)" & sce$strain != "Cx3"]
sce_c$fl <- droplevels(factor(sce_c$fl))
```

Set the genes for the heatmap
```{r}
# The top 30 MG(A2) DEGs, resident MG marker Serpine2 and infiltrating marker Gpnmb
input <- c("Serpine2", "Gpnmb", rownames(markers[[2]])[1:30]) 
```

Plot the heatmap
```{r}
library(scater)

#plot and save the heatmap
plotHeatmap(sce_c, features=input, exprs_values="logcounts", cellheight=5,
            zlim=NULL, center=FALSE, symmetric=FALSE, cluster_cols=F, cluster_rows=T,
            colour_columns_by=c("fl"), columns=order(sce_c$fl), fontsize_row=5, fontsize_col=0.1, treeheight_col=10, file=dirH, fontsize=7)
```
