---
title: "Figure 2 + S. Figure 3"
author: "Regan"
date: "16/05/2020"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

This script will generate the scRNA-seq data figures in figure 2 and extended data figure 3 from the dataset that has been pre-processed with scripts 1-6.

Load the sce
```{r}
rm(list=ls())
sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")
```

# Fig 2a, MG downregulate canonical marker genes. Seurat dot plot
```{r}
library(Seurat)
library(ggplot2)


# extract the MG and convert the sce to a Seurat object
seur <- as.Seurat(sce[,sce$cell_type == "MG"], data="logcounts", assay="logcounts")

MG_genes <- c("Olfml3","Sparc", "Trem2", "C1qa", "C1qb",  "Ctss","Csf1r", "Hexb","Gpr34", "P2ry12","Plxdc2", "Sall1", "Siglech","Cx3cr1","Tmem119","Serpine2")

DotPlot(seur,
        features=MG_genes,
        group.by = "stage",
        cols = "RdBu",
        scale =F,
        assay="logcounts") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

ggsave("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 2a - MG markers.pdf", width = 8, height = 2.5)
```


# Fig 2b, UMAP displaying Apoe vs Trem2 expression
- using Seurat to plot the expression of two genes one one UMAP

Load the sce
```{r}
rm(list=ls())
sce_u <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

dirU <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 2b "
```

Plot the UMAP
```{r}
library(Seurat)
# convert the sce to a seurat object
seur <- as.Seurat(sce_u)

# select the genes to plot
x <- "Apoe"
y <- "Trem2"

# generate the UMAPs
list <- FeaturePlot(object=seur,
                    features=c(x, y), 
                    cols=c("grey60", "purple", "green"), 
                    reduction="UMAP",
                    blend=TRUE,
                    combine=F,
                    blend.threshold=0,
                    pt.size=1)


# convert the UMAP dots to a single image (easier handling in Illustrator)
AugmentPlot(plot=list[[3]] + NoLegend() + labs(title=NULL))  

ggsave(paste0(dirU,  x, " vs ", y, " .pdf"), scale=0.75, width=7, height=8,units=c("in", "cm", "mm"), dpi=300, limitsize=TRUE)

# plot and save the legend
list[[4]]+NoLegend()
ggsave(paste0(dirU, x, " vs ", y, " r-b legend.pdf"), scale=0.75, width=8, height=8,units=c("in", "cm", "mm"), dpi=300, limitsize=TRUE)
```

# Fig 2c, d and S.Fig 3 Boxplots of genes of interest across SCI timepoints

Load the sce and directory
```{r}
rm(list=ls())
sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")
sce$cell_type <- factor(sce$cell_type)

dirB <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/S. Fig2/"
```

Set the gene lists of interest
```{r}
MG <- c("P2ry12","Plxdc2", "Sall1", "Siglech", "Sparc","Serpine2")
names(MG) <- "MG"

MG_MCd <- c("Tmem119", "Gpr34", "Olfml3")
names(MG_MCd) <- "MG_MCd"

Myel <- c("Aif1", "Hexb", "Fcrls","C1qa", "Trem2")
names(Myel) <- "Myel"

YFP <- c("S100a11", "Ms4a7", "Chil3", "Mgst1", "Ccl7", "Arg1")
names(YFP) <- "Infiltrating"

Fig2 <- c("Apoe", "Tyrobp")
names(Fig2) <- "Fig2"
```

Generate the boxplots for the genes of interest (Welch t-Test)
```{r}
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
# set the genes of interest
genes <- Fig2

x <- sce$stage
col <- sce$cell_type

for (i in genes) {
  
  logs <- as.numeric(logcounts(sce[i]))
  df <- data.frame(Logs=logs, x=x, colour=col)
  
  #add empty slots for HC to keep the spacing
  df.a <- data.frame(Logs=c(0,0,0, 0, 0), x=c("HC", "HC", "HC", "10", "21"), colour=c("MCd", "DC", "NP", "NP", "NP"))
  df <- rbind(df, df.a)
  comparisons = list(c("HC", "1"), c("HC", "2"), c("HC", "3"), c("HC", "10"), c("HC", "21"))
  
  if (names(genes) == "MG" | names(genes) == "MG_MCd") {
    sym <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns"))
    statP <- stat_compare_means(method = "t.test", comparisons = comparisons, symnum.args = sym)
  } else { 
    statP <- NULL
  }
  ggplot(data=df, aes(y=Logs, x=x, fill=colour)) + geom_boxplot() + ggtitle(i) + theme(plot.title=element_text(hjust=0.5), legend.position="none") + ylab("Logcounts") + xlab("SCI Stage (days)") + scale_fill_brewer(palette="Accent", direction=1) + statP
  
  ggsave(paste0(dirB,"boxplot_", i, ".pdf"), scale=.75, height=5, width =5)
}

# Statistics for extended data Table 3, complementary to extended data fig 2
if (names(genes) == "MG" | names(genes) == "MG_MCd" | names(genes) == "Myel") {
  # Perform pairwiseTTests and multiple testing correction via Benjamini-Hochberg for the genes of interest 
  # set the cell type
  sce_CT <- sce[,sce$cell_type == "MG"]
  
  # calculate the adjusted pval
  library(scran)
  out <- pairwiseTTests(logcounts(sce_CT),
                        groups=sce_CT$stage,
                        direction="down")
  # extract the pairs of interest (ie HC vs each timepoint)
  pairs <- c(6, 11, 16, 21, 26)
  
  # extract just the adjusted p values for each gene 
  stats <- out$statistics[[6]][genes,]$FDR
  stats_a <- out$statistics[[6]][genes,]
  
  for (i in pairs[-1]) {
    add <- out$statistics[[i]][genes,]$FDR
    add_a <- out$statistics[[i]][genes,]
    
    stats_a <- cbind(stats_a, add_a)
    stats <- cbind(stats, add)
  }
  colnames(stats) <- out$pairs$first[pairs]
  rownames(stats) <- genes
  
  # save and add to plot in illustrator
  write.csv(stats, paste0(dirB, "stats", names(genes)[1], " x vs HC.csv"))
  write.csv(stats_a, paste0(dirB, "stats-all ", names(genes)[1], "  x vs HC.csv"))
}
```


# Fig 2d boxplot of DAM1 vs DAM2 genes by cell type over time

Load the sce and directory
```{r}
rm(list=ls())
sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")
sce$cell_type <- factor(sce$cell_type)

dirB <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 2d/"
```

Load the function for preparing gene lists from source files
```{r}
source("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/scRNAseq Workflow/Regan et al 202X/Sources/Fxn - Prep Genes.R")
```

Establish the DAM gene lists (from Keren-Shaul et al., 2017; Figure 6)
```{r}
# DAM phase 1 (Trem independent)
DAM1 <- c("Tyrobp, Ctsb, Ctsd, Apoe, B2m, Fth1, Lyz2") 

# DAM phase 2 (Trem dependent)
DAM2 <- c("Trem2, Axl, Cst7, Ctsl, Lpl, Cd9, Csf1, Ccl6, Itgax, Clec7a, Lilrb4a, Timp2")  

# separate the string
DAM1 <- prepG(DAM1, sep = ", ")
DAM2 <- prepG(DAM2, sep = ", ")
```

Calculate the per cell metrics for the DAM genes
```{r}
library(scater)

#sce <- calculateQCMetrics(sce, feature_controls=list(DAM1=which(rownames(sce) %in% DAM1), DAM2=which(rownames(sce) %in% DAM2)))
sce$logMeanDAM2 <- colMeans(logcounts(sce[DAM2]))
sce$logMeanDAM1 <- colMeans(logcounts(sce[DAM1]))
```

Generate the boxplots for the genes of interest
```{r}
library(RColorBrewer)

# set the x values
x <- sce$stage

# set the values to colour the plot by
col <- sce$cell_type 

# set the per cell metric level
sublevel <- "logMean" #total_features_by_counts_ log10_total_counts_ pct_counts_

# the gene lists names for designating which per cell metrics, and for labelling plots
names <- c("DAM1", "DAM2")

# generate the plots for each gene list
for (i in names) {
  
  # build the data frame for plotting
  sl <- paste0(sublevel, i )
  logs <- colData(sce)[sl][,1]
  
  df <- data.frame(Logs=logs, x=x, colour=col)
  
  # add empty slots for HC to keep the spacing
  df.a <- data.frame(Logs=c(0,0,0, 0), x=c("HC", "HC", "HC", "10"), colour=c("MCd", "DC", "NP", "NP"))
  df <- rbind(df, df.a)
  
  # plot
  ggplot(data=df, aes(y=Logs, x=x, fill=colour)) + geom_boxplot() + ggtitle(i) + theme(plot.title=element_text(hjust=0.5), legend.position="none") + ylab(sl) + xlab("Day Post-SCI") + scale_fill_brewer(palette="Accent", direction=1) + ggtitle(i)
  
  # save
  ggsave(paste0(dirB,"boxplot-", i, sublevel, "MG FIG6 list by clust.pdf"), scale=.75, height=5, width=5)
}
```

# Statistics for Extended Data Table 2, complementary to fig 2d
```{r}
genes <- DAM2

# select only the subacute and chronic cells
sce_CT <- sce[,sce$stage == "10" | sce$stage == "21"]

# calculate the adjusted pval
library(scran)
out <- pairwiseTTests(logcounts(sce_CT),
                      groups=sce_CT$cell_type,
                      direction="up")

# extract the genes and comparison (1=MG vs MCd) of interest
stats_a <- as.matrix(out$statistics[[1]][genes,])

# extract just the adjusted p values for each gene 
stats <- as.matrix(out$statistics[[1]][genes,]$FDR)

colnames(stats) <- out$pairs$first[1]
rownames(stats) <- rownames(stats_a) <- genes


# save and add to plot in illustrator
write.csv(stats, paste0(dirB, "stats D10-21 MG vs MCd DAM2.csv"))
write.csv(stats_a, paste0(dirB, "stats-all D10-21 MG vs MCd DAM2.csv"))
```


