---
title: "Figure 1"
author: "Regan"
date: "16/05/2020"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

This script will generate the scRNA-seq data figures in figure 1 from the dataset that has been pre-processed with scripts 1-6.


# Fig 1b,c - UMAPs of the dataset separated by strain and fate mapping, coloured by the day collected post-SCI.

Load the sce, directory, and temporal colour palette
```{r}
sce_all <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

dirU <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 1b/"

# Set the colour palette (red to purple) for data related to timepoints post-SCI
temp <- c("#D53E4F", "#FF8D59","#E6F598","#1A9850","#3288BD", "#542788")
```

Plot the UMAP using the mouse strain/fluorophore of interest and grey out the cells not of interest
```{r}
library(scater)
library(RColorBrewer)
library(Seurat)
sce <- sce_all

# name for labelling the UMAP
name <- "grey by fl"

# choose how to colour and shape the UMAP
colours <- c(temp,"grey95")

# choose the level to determine the grey
subset <- "fl"

# choose the level to colour the non-grey cells
col.name <- "stage"

# automatically set parameters
level <- colData(sce)[subset][,1]

# load the sce and generate the UMAP for each timepoint
for (j in levels(level)) {
  sce <- sce_all
  
  #set the colour category for the grey
  grey <- level != j
  
  # rename levels to include "grey" for those not of interest
  col <- colData(sce)[col.name][,1] # set the level to colour the non-grey cells
  sce$grey <- factor(col, levels=c(levels(col), "grey"))
  sce$grey[grey] <- "grey"
  
  #sort the cells in order to have the grey cells last (and therefore underneath)
  sce <- sce[,rev(order(colnames(sce)[sce$grey]))]
  
  # generate the UMAP for the timepoint
  p <- plotUMAP(sce, colour_by="grey",by_exprs_values="logcounts", point_size = 1) +  ggtitle(paste0(j))  + theme(plot.title=element_text(size=rel(2)))   + geom_point(aes(fill=colData(sce)$grey), colour="grey95", pch=21, stroke = .1, size =1.5, alpha = 0.6) + scale_fill_manual(values=colours, drop=F)+ theme(legend.position="none")
  
  # convert the UMAP dots to a single image (easier handling in Illustrator)
  p <- AugmentPlot(plot=p, dpi=600)  
  
  # name and save the plot to combine with the other timepoints  
  ggsave(p, filename=paste0(name," ", j, ".pdf"), path=dirU, scale=0.7, width=3.5, height=4, dpi=300, limitsize=TRUE) 
}

# combine the UMAPs from all timepoints and save

```

# Fig 1d Cell Type Markers

Load the sce and directory
```{r}
rm(list=ls())

sce <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

dirT <- "/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 1d/"
```

Define the cell type markers
```{r}
MG <-c("Serpine2", "P2ry12", "Siglech", "Slc2a5", "Plxdc2", "Sall1", "Sparc")
NP <- c("S100a8", "S100a9", "Lcn2", "Dedd2", "Ly6g")
MC <- c("Ccl7", "Ecm1", "Mgst1", "Arg1", "Ms4a7")
DC <- c("Cd74", "H2-Eb1", "H2-Aa", "Flt3", "Zbtb46" )

CT <- list(MG=MG, NP=NP, MC=MC, DC=DC)
```

Colour the UMAP by marker genes
```{r}
library(scater)
set.seed(100)

# choose how to colour and shape the tSNE
c.by <- "logMean"
s.by <- NULL 
exp <- "logcounts"

x <- 0
for (i in CT) {
  x <- x+1
  # set the cell type
  names <- names(CT)[x]
  genes <- i
  
  # combine the genes into a single string for labelling the plot
  g_names <- NULL
  for(j in genes){
    g_names <- paste(g_names, j)
  }
  
  # calculate the mean expression value (from log-transformed counts) for the genes of interest
  sce$logMean <- colMeans(logcounts(sce[genes,]))
  
  # plot the UMAP
  p <- plotUMAP(sce, colour_by=c.by, shape_by=s.by, by_exprs_values=exp) +  ggtitle(paste0(g_names," P=", metadata(sce)$perplexity)) 
  
  # convert the UMAP dots to a single image (easier handling in Illustrator)
  p <- AugmentPlot(plot=p, dpi=600)  
  
  #save the UMAP and adjust the size as needed
  ggsave(p, filename=paste0("Unique DEGs logMean -", names, ".pdf"), path=dirT, scale=1, width=9, height=9, dpi=300, limitsize=TRUE) 
}

```

# Fig 1e (left) Area composition plot, cell type vs timepoint
- using the Cx3 mouse only, because it is an unbiased sampling of Cx3cr1+ cells compared to the Cremato, which enriches for either infiltrating or resident Cx3cr1+ cells

Load the sce
```{r}
rm(list=ls())
sce_all <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")

#examine only Cx3cr1 mice
sce_comp <- sce_all[, sce_all$fl == "Cx3"]

#reorder the cell types for aesthetics 
sce_comp$cell_type <- factor(sce_comp$cell_type, levels=c("NP","DC","MCd","MG"))
```

Generate the area plot (using sce_comp)
```{r}
library(tibble)
library(ggplot2)
library(plyr)


# set HC to 0
sce_comp$stage <- factor(sce_comp$stage, levels = c("HC","0", "1", "2", "3", "10", "21"))
sce_comp$stage[sce_comp$stage == "HC"] <- 0

# set the x and y values for the plot
y <- sce_comp$cell_type
x <- droplevels(factor(sce_comp$stage))

# create the df containing the x and y values
df <- as.data.frame(table(y, x))
colnames(df) <- c("y", "x", "Freq")

# create data
time <- as.numeric(levels(df$x)[df$x]) # x Axis
value <- as.numeric(df$Freq)             # y Axis
group <- df$y       # group, one shape per group
data <- data.frame(time, value, group)


# plot
# Compute percentages with dplyr
library(dplyr)
data <- data  %>%
  group_by(time, group) %>%
  summarise(n = sum(value)) %>%
  mutate(percentage = n / sum(n))

# Plot
# Plot
ggplot(data, aes(x=time, y=percentage, fill=group)) + 
  geom_area(alpha=0.7 , size=0.5, colour="black") + ylab("Fraction of Cells Per Cell Type") + scale_x_continuous(name ="Day Post-SCI", breaks=c(0,1,2, 3, 10, 21), labels=c("0"="HC","1"="1","2"="2", "3"="3", "10"="10", "21"="21"))+ scale_fill_brewer(palette="Accent", direction=-1)


ggsave("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 1e/Composition Area Plot Cell Type.pdf", height=5, width=4)
```

# Fig 1e (right) UMAPs by timepoint coloured by cell type

Load the sce and directory
```{r}
library(scater)

# load the sce
sce_all <- readRDS("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/sce.RDS")
```

Generate UMAPs for each timepoint with greyed out background cells
```{r}
library(scater)
library(RColorBrewer)
library(Seurat)

# set the days for subsetting the sce
days <- c("HC","1", "2", "3", "10", "21") 

# choose how to colour and shape the UMAP
c.by <- "grey"
s.by <- NULL 
exp <- "logcounts"
colours <- c(brewer.pal(nlevels(sce_all$cell_type), "Accent"),"grey95")

# name for labelling the UMAP
name <- c.by

# load the sce and generate the UMAP for each timepoint
for (j in days) {
  sce <- sce_all
  #set the colour category for the UMAP
  col.name <- "cell_type"
  col <- sce$cell_type #sce$cells.condition 
  
  #set the colour category for the grey
  category <- j
  subset <- sce$stage
  grey <- subset != category
  
  # rename levels to include "grey" for those not of interest
  sce$grey <- factor(col, levels=c( levels(col), "grey"))
  sce$grey[grey] <- "grey"
  
  library(dplyr)
  #sort the cells in order to have the grey cells last (and therefore underneath)
  sce <- sce[, sce$grey == j] %>% cbind(sce[, sce$grey == "grey"])

  
  # generate the UMAP for the timepoint
  p <- plotUMAP(sce, colour_by=c.by, shape_by=s.by, by_exprs_values=exp, point_size = 1) +  ggtitle(paste0(j))  + theme(plot.title=element_text(size=rel(2)))  + geom_point(aes(fill=sce$grey),colour= "grey95",  pch=21, stroke = .1, size =1.5) + scale_fill_manual(values=colours, drop=F) + theme(legend.position="none") # #+ xlim(-12.5,7.5) + ylim(-12.5, 12.5)
  
    # convert the UMAP dots to a single image (easier handling in Illustrator)
  p <- AugmentPlot(plot=p, dpi=600)  
  
  p 
  
  # save the plot 
  ggsave(paste0("/Users/reganhamel/Dropbox (Cambridge University)/Lab/Regan/SCI scRNA-seq Project/Cx3cr1 Data/Universal Workflow/Cell Ranger 3.1.0/SCI/May 2020/BC/Hamel et al 202X/Fig 1e/grey CT D",j,".pdf"), width=9, height=9)
}


```
